---
title: "Effect of Car Design Aspects on Fuel Economy"
author: Jonathan Owen
date: July 26th 2015
output: 
  pdf_document:
    latex_engine: xelatex
mainfont: Georgia
fontsize: 10pt
geometry: margin = 0.75in
---
```{r, echo = FALSE, message = FALSE}
library(ggplot2)
library(gridExtra)
```
 

## Executive Summary
The fuel economy of a car should be related to some of its design aspects. The 
following study uses reported values of fuel economy and these design aspects to 
establish relationships in the form of linear regression models.  Horsepower,
weight, and time to travel $\frac{1}{4}$ mile are found to be the most useful 
predictors of fuel consumption.  

## Data
The `mtcars` dataset contains observations of fuel economy and 10 design aspects 
reported in _Motor Trend_ magazine for 32 car models available in the United 
States between 1973 and 1974. The variables are described in the table below.

------ | ----------------------------- | -----  | ----------------------------- | -----  | ----------------------------- |  
`mpg`  | fuel economy, miles/US gallon | `drat` | rear axle ratio               | `am`   | transmission                  |              
`cyl`  | number of cylinders           | `wt`   | weight, 1000 lb               | `gear` | number of forward gears                |                               |                
`disp` | displacement, cubic inches    | `qsec` | 1/4 mile time, seconds        | `carb` | number of carburettors               |                
`hp`   | gross horsepower              | `vs`   | V or straight cylinder        |        |   

Exploratory data analysis identified several possible predictors as well 
as issues prior to constructing parsimonious models for fuel economy. A panel of 
scatterplots and correlations for all variable pairs is included as Figure 1 in 
the appendix. For reasons explained in the **Initial Model** section, it is 
easier to understand the model in terms of fuel consumption `gpm` in US gallons 
per mile rather than fuel economy `mpg`.  The two quantities are related by each 
being the reciprocal of the other. Plots of fuel consumption against the 10 design 
aspect variables show strong negative correlations (absolute values 0.78 - 0.87) 
with displacement, horsepower, and weight. It is also clear that these variables 
are positively correlated with each other (0.66 - 0.89), which may cause 
problems if they are all are used in the model. The number of cylinders falls 
into this same group, but is not an initial choice for the model because it is 
not continuous and is correlated with displacement (0.90) as would be expected.


## Initial Model
The exploratory data analysis can be supplemented with some theory to create an 
initial model.
The work done by a moving car during a time period $\Delta t$, such as `qsec` in 
the dataset, is $P(t)\Delta t$ where $P(t)$ is the power at time $t$. Although 
the exact form of $P(t)$ is not reported for these cars, it should be 
proportional to the gross horsepower, which *is* available. The work done should 
be proportional to both `hp` and `qsec`.  
The energy required to do this work comes from the fuel and is proportional to 
the volume of fuel used. Not all of this energy goes into driving the car. 
There are losses that can be grouped into an efficiency factor $\eta$, which 
should depend on some of the design aspects.  The following relationship 
should be true for a car travelling $\frac{1}{4}$ mile  
$$\frac{\eta}{4\cdot\mathtt{mpg}} \propto \mathtt{hp}\cdot\mathtt{qsec}$$ 
or $$\frac{\eta\cdot\mathtt{gpm}}{4} \propto \mathtt{hp}\cdot\mathtt{qsec}$$  

The inital model is a fit of `gpm` against the interaction `hp:qsec`. 
$$\mathtt{gpm}_{i} = \beta_{0} + \beta_{1}\mathtt{hp}_{i}\cdot\mathtt{qsec}_{i} + \epsilon_{i}$$
The intercept $\beta_{0}$ is not significant ($p = 0.32$). $R^2$ increases from 
0.69 to 0.96 when the intercept is removed from the model.

```{r, echo = FALSE}
save_cols <- names(mtcars)
c <- mtcars
nice_names <- c("cylinders", "displacement", "horsepower", 
                "rear axle\nratio", "weight", "1/4 mile\ntime", 
                "cylinder\nconfiguration", "transmission", "gears", 
                "carburettors", "fuel\nconsumption")
c$gpm <- 1/c$mpg
c$idi <- 1/c$disp
c$ihp <- 1/c$hp
c$iqs <- 1/c$qsec
c$iwt <- 1/c$wt


gpm1 <- lm(data = c, gpm ~ hp:qsec)
gpm0 <- lm(data = c, gpm ~ hp:qsec -1)
```

## Model Improvement by Residuals Adjustment
Residuals adjustment were used to find any additional predictors---ones that are 
rolled into the efficiency factor $\eta$. Residuals from fitting `gpm` and the 
remaining car design aspects using `hp:qsec` are plotted against each 
other.  These are included in Figure 2.  High correlation in the residuals plots 
identifies predictors that will best fit variation that remains from the initial 
model. Of the unused predictors, `wt` has the highest correlation (0.81) with 
`gpm` after removing the effects of `hp` and `qsec`.  A second model that
included `wt` as a predictor 
$$\mathtt{gpm}_{i} = \beta_{1}\mathtt{hp}_{i}\cdot\mathtt{qsec}_{i} + \beta_{2}\mathtt{wt}_{i} + \epsilon_{i}$$
```{r, echo = FALSE}
gpm2 <- lm(data = c, gpm ~ hp:qsec + wt -1)

```
produces a model with significant coefficients $\beta_{1} = 5.92 \times 10^{-6}$ 
and  $\beta_{2} = 1.20 \times 10^{-2}$ and $R^2$ of 0.99. 

## ANOVA and Residuals Analysis
Analysis of variance ANOVA on the nested models indicates that reduction of the 
sum of the residual squared from 0.0042 to 0.0041 by including `wt` is 
significant ($p = 1.3 \times 10^{-8}$).  

Lastly, four plots of the residuals can be used to assess the fit. These are
included in Figure 3. 
In the plot of residuals vs. fitted values points should be randomly distributed 
around a horizontal line corresponding to zero residual. This is approximately 
the case in Figure 2 although there is some deviation at higher fitted values. 
Similar behavior is expected for standardized residuals vs. fitted values and 
this is also approximately the case for the model.  Of note is that all the 
standardized residuals appear to be within 2 standard deviations of zero.

If the residuals are normal they should lie close to a diagonal line in the Q-Q 
plot. Again, this is mainly true but there is some deviation at lower values.  

In the plot of residuals vs. leverage, there are some concerns about high 
leverage points, such as the Maserati Bora; however, the standardized residuals 
for this point does show it has a large effect despite the potential leverage. 
Of greater concern is the location of Chrysler Imperial point in relation to the 
Cook's distance 0.5 contour. Cook's distance measure the effect of deleting a 
point on the model.  Higher Cook's distances correspond to outliers (high 
residuals) or high leverage.  Ideally, the Chrysler Imperial requires further
examination.

Overall, the residuals analyses suggest that, although the model has a high 
$R^2$ value, there may be missing predictors because the distribution of 
residuals does not appear entirely random.  However, testing with other 
variables in the mtcars dataset did nothing to improve the appearance of 
residuals plots. Additional predictors that could improve the fit do not seem 
to have been included in the dataset.

## Intepretation of the coefficients and uncertainty in the model
Coefficient $\beta_{1}$ implies that for an increase of 1 hp the expected fuel 
consumption increases by $5.92 \times 10^{-6}$ US gallons per mile if all 
other variables are unchanged.  The same change occurs for each 1 second extra 
on the $\frac{1}{4}$ mile time.
Coefficient $\beta_{2}$ implies that for each 1000lb increase in the weight of 
the car, the expected fuel consumption increases by $1.20 \times 10^{-2}$ US 
gallons per mile.
Although these are the expected values, there is uncertainty in these estimates.
The 95% confidence intervals are $1.99 \times 10^{-6}$ to $9.86 \times 10^{-6}$ 
for $\beta_{1}$ and $0.89 \times 10^{-2}$  to $1.52 \times 10^{-2}$ for 
$\beta_{2}$. Separate prediction intervals would need to be calculated for
predictions using the model with new data.  

## Effect of transmission type
Unfortunately, the model does not directly address the effect of transmission
type. As noted in the exploratory data analysis, `am` is correlated with `wt` 
(0.69) as well as other variables of interest.  In the dataset the 
weight range for manual transmission cars is 1,500 - 3,500 lbs and for 
automatics is 2,500 - 4,500 lbs.  The overlap of the two ranges doesn't provide 
enough data points to try to separate the effect of transmission from weight. 
Also, outside of the data, the effect of automatic transmissions is consider to
be primarily due to the increased weight. So even though the transmission isn't
include the difference between a manual and automatic transmission versions of 
otherwise identical cars could be predicted using their weights.
  

\newpage

## Appendix

The color of the points indicates transmission type in all plots. 
**\textcolor{red}{Red points}** are automatic. 
**\textcolor{blue}{Blue points}** are manual.  
&nbsp;  

**Figure 1** Scatterplot and correlation of variable pairs from the `mtcars` dataset  

```{r, fig.height = 9, fig.width = 9, echo = FALSE}
panel.pearson <- function(x, y, ...) {
horizontal <- (par("usr")[1] + par("usr")[2]) / 2; 
vertical <- (par("usr")[3] + par("usr")[4]) / 2; 
text(horizontal, vertical, format(abs(cor(x,y)), digits=2)) 
}

pairs(c[,2:12], gap = 0, pch = 16, pty = "m",
      cex = 0.75, cex.labels = 1, cex.axis = 1, 
      tcl = -0.25,  mgp =c(0, 0.3 , 0), adj = 0.5,
      col = rgb(red = (c$am == 0), 
                green = 0, 
                blue = (c$am == 1), 
                alpha = 0.5), 
      labels = nice_names,
      upper.panel = panel.pearson)
```

\newpage

**Figure 2** Residuals of `gpm` vs. other variables after fitting with `hp:qsec`

```{r, echo = FALSE}
rgpm <- resid(gpm0)
cyl1 <- lm(data = c, cyl ~ hp:qsec - 1)
rcyl <- resid(cyl1) 
rrcyl <- lm(data = c, rgpm ~ rcyl)
gpm_cyl <- round(cor(rcyl, rgpm),2)
pcyl <- ggplot(data = c, aes(x = rcyl, y = rgpm)) 
pcyl <- pcyl + geom_point(col = rgb(red = (c$am == 0),
                                  green = 0, 
                                  blue = (c$am == 1), 
                                  alpha = 0.5),
                        shape = 16, size = 2) 
pcyl <- pcyl + geom_abline(intercept = coef(rrcyl)[1], slope = coef(rrcyl)[2])
pcyl <- pcyl + theme_bw() + ggtitle(paste("cylinders", gpm_cyl, sep = "   ")) + 
        theme(axis.title = element_blank(), 
        plot.title = element_text(size = 10, vjust = 0.25))
```

```{r, echo = FALSE}
disp1 <- lm(data = c, disp ~ hp:qsec - 1)
rdisp <- resid(disp1) 
rrdisp <- lm(data = c, rgpm ~ rdisp)
gpm_disp <- round(cor(rdisp, rgpm),2)
pdisp <- ggplot(data = c, aes(x = rdisp, y = rgpm)) 
pdisp <- pdisp + geom_point(col = rgb(red = (c$am == 0),
                                  green = 0, 
                                  blue = (c$am == 1), 
                                  alpha = 0.5),
                        shape = 16, size = 2) 
pdisp <- pdisp + geom_abline(intercept = coef(rrdisp)[1], slope = coef(rrdisp)[2])
pdisp <- pdisp + theme_bw() + ggtitle(paste("displacement", gpm_disp, sep = "   ")) + 
         theme(axis.title = element_blank(), 
         plot.title = element_text(size = 10, vjust = 0.25))
```

```{r, echo = FALSE}
drat1 <- lm(data = c, drat ~ hp:qsec - 1)
rdrat <- resid(drat1) 
rrdrat <- lm(data = c, rgpm ~ rdrat)
gpm_drat <- round(cor(rdrat, rgpm),2)
pdrat <- ggplot(data = c, aes(x = rdrat, y = rgpm)) 
pdrat <- pdrat + geom_point(col = rgb(red = (c$am == 0),
                                  green = 0, 
                                  blue = (c$am == 1), 
                                  alpha = 0.5),
                        shape = 16, size = 2) 
pdrat <- pdrat + geom_abline(intercept = coef(rrdrat)[1], slope = coef(rrdrat)[2])
pdrat <- pdrat + theme_bw() + ggtitle(paste("rear axle ratio", gpm_drat, sep = "   ")) +
         theme(axis.title = element_blank(), 
         plot.title = element_text(size = 10, vjust = 0.25))
```

```{r, echo = FALSE}
wt1 <- lm(data = c, wt ~ hp:qsec - 1)
rwt <- resid(wt1) 
rrwt <- lm(data = c, rgpm ~ rwt)
gpm_wt <- round(cor(rwt, rgpm),2)
pwt <- ggplot(data = c, aes(x = rwt, y = rgpm)) 
pwt <- pwt + geom_point(col = rgb(red = (c$am == 0),
                                  green = 0, 
                                  blue = (c$am == 1), 
                                  alpha = 0.5),
                        shape = 16, size = 2) 
pwt <- pwt + geom_abline(intercept = coef(rrwt)[1], slope = coef(rrwt)[2])
pwt <- pwt + theme_bw() + ggtitle(paste("weight", gpm_wt, sep = "   ")) +
       theme(axis.title = element_blank(), 
       plot.title = element_text(size = 10, vjust = 0.25))
```

```{r, echo = FALSE}
vs1 <- lm(data = c, vs ~ hp:qsec - 1)
rvs <- resid(vs1) 
rrvs <- lm(data = c, rgpm ~ rvs)
gpm_vs <- round(cor(rvs, rgpm),2)
pvs <- ggplot(data = c, aes(x = rvs, y = rgpm)) 
pvs <- pvs + geom_point(col = rgb(red = (c$am == 0),
                                  green = 0, 
                                  blue = (c$am == 1), 
                                  alpha = 0.5),
                        shape = 16, size = 2) 
pvs <- pvs + geom_abline(intercept = coef(rrvs)[1], slope = coef(rrvs)[2])
pvs <- pvs + theme_bw() + ggtitle(paste("cylinder configuration", gpm_vs, sep = "   ")) +
       theme(axis.title = element_blank(), 
       plot.title = element_text(size = 10, vjust = 0.25))
```

```{r, echo = FALSE}
am1 <- lm(data = c, am ~ hp:qsec - 1)
ram <- resid(am1) 
rram <- lm(data = c, rgpm ~ ram)
gpm_am <- round(cor(ram, rgpm),2)
pam <- ggplot(data = c, aes(x = ram, y = rgpm)) 
pam <- pam + geom_point(col = rgb(red = (c$am == 0),
                                  green = 0, 
                                  blue = (c$am == 1), 
                                  alpha = 0.5),
                        shape = 16, size = 2) 
pam <- pam + geom_abline(intercept = coef(rram)[1], slope = coef(rram)[2])
pam <- pam + theme_bw() + ggtitle(paste("transmission", gpm_am, sep = "   ")) +
       theme(axis.title = element_blank(), 
       plot.title = element_text(size = 10, vjust = 0.25))
```

```{r, echo = FALSE}
gear1 <- lm(data = c, gear ~ hp:qsec - 1)
rgear <- resid(gear1) 
rrgear <- lm(data = c, rgpm ~ rgear)
gpm_gear <- round(cor(rgear, rgpm),2)
pgear <- ggplot(data = c, aes(x = rgear, y = rgpm)) 
pgear <- pgear + geom_point(col = rgb(red = (c$am == 0),
                                  green = 0, 
                                  blue = (c$am == 1), 
                                  alpha = 0.5),
                        shape = 16, size = 2) 
pgear <- pgear + geom_abline(intercept = coef(rrgear)[1], slope = coef(rrgear)[2])
pgear <- pgear + theme_bw() + ggtitle(paste("gears", gpm_gear, sep = "   ")) +
         theme(axis.title = element_blank(), 
         plot.title = element_text(size = 10, vjust = 0.25))
```

```{r, echo = FALSE}
carb1 <- lm(data = c, carb ~ hp:qsec - 1)
rcarb <- resid(carb1) 
rrcarb <- lm(data = c, rgpm ~ rcarb)
gpm_carb <- round(cor(rcarb, rgpm),2)
pcarb <- ggplot(data = c, aes(x = rcarb, y = rgpm)) 
pcarb <- pcarb + geom_point(col = rgb(red = (c$am == 0),
                                  green = 0, 
                                  blue = (c$am == 1), 
                                  alpha = 0.5),
                        shape = 16, size = 2) 
pcarb <- pcarb + geom_abline(intercept = coef(rrcarb)[1], slope = coef(rrcarb)[2])
pcarb <- pcarb + theme_bw() + ggtitle(paste("carburettors", gpm_carb, sep = "   ")) +
         theme(axis.title = element_blank(), 
         plot.title = element_text(size = 10, vjust = 0.25))
```

```{r, fig.height = 9, echo = FALSE}
grid.arrange(pcyl, pdisp, pdrat, pwt, pvs, pam, pgear, pcarb, ncol = 2)
```

\newpage

**Figure 3** Residuals analysis of the final model

```{r, fig.height = 3.3, fig.width = 3.3, echo = FALSE}
plot(gpm2, pch = 16)
```


